Supplementary Appendix

### 2025-07-29 paper/preprint/appendix.tex
final pass-through done

\begin{itemize}
\tightlist
\item
  \(\mathcal{Y}\): The output domain.
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(y^-\): The non-target class and also the index of non-the target
  class.
\item
  \(\mathcal{X}\): The input domain.
\item
  \(\mathbf{x}\): a single training sample.
\item
  \(\mathbf{x}^\prime\): a counterfactual.
\item
  \(t=1,...,T\): Step indicator for counterfactual search iterations.
\item
  \(\mathbf{x}^\prime_{t \leq T}\): a nascent counterfactual, defined as
  a counterfactual that has not yet converged.
\item
  \(\mathbf{x}^\prime_{T}\): a mature counterfactual, defined as a
  counterfactual that has either passed a threshold probability or
  exhausted all \(T\) steps.
\item
  \(\mathbf{x}^+\): a training sample in the target class
  (ground-truth).
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\item
  \(\mathbf{M}(\cdot)\): linear predictions (logits) of the classifier.
\end{itemize}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, colback=white]

@@

For all of our counterfactual evaluations, we proceed as follows: for
each dataset we run \(J\) bootstrap rounds (``No.~Runs'') to account for
stochasticity (Note~\ref{nte-eval-default}); for each bootstrap round,
we randomly draw factual and target pairs; then, for each model, we draw
samples from the test set (with replacement) for which the model
predicts the randomly chosen factual class; finally, we generate
multiple counterfactuals (``No.~Counterfactuals'') and evaluate the
outcomes (Note~\ref{nte-eval-default}). This is in line with standard
practice in the related literature on CE. For our final results
presented in the main paper, we rely on held-out test sets for
evaluation. For tuning purposes we rely on training and/or validation
sets.

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, colback=white]

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, colback=white]

@@

>>>>

\begin{itemize}
\tightlist
\item
  \(\mathcal{Y}\): The output domain.
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(y^-\): The non-target class and also the index of non-the target
  class.
\item
  \(\mathcal{X}\): The input domain.
\item
  \(\mathbf{x}\): a single training sample.
\item
  \(\mathbf{x}^\prime\): a counterfactual.
\item
  \(t=1,...,T\): Step indicator for counterfactual search iterations.
\item
  \(\mathbf{x}_{\text{AE}}^\prime\): a nascent counterfactual, defined
  as a counterfactual that has not yet converged.
\item
  \(\mathbf{x}_{\text{CE}}^\prime\): a mature counterfactual, defined as
  a counterfactual that has either passed a threshold probability or
  exhausted all \(T\) steps.
\item
  \(\mathbf{x}^+\): a training sample in the target class
  (ground-truth).
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\item
  \(\mathbf{M}(\cdot)\): linear predictions (logits) of the classifier.
\end{itemize}

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

For all of our counterfactual evaluations, we proceed as follows: for
each dataset we run \(J\) bootstrap rounds (``No.~Runs'') to account for
stochasticity (Note~\ref{nte-eval-default}); for each bootstrap round,
we randomly draw factual and target pairs; then, for each model, we draw
samples from the test set (with replacement) for which the model
predicts the randomly chosen factual class; finally, we generate
multiple counterfactuals (``No.~Counterfactuals'') and evaluate the
outcomes (Note~\ref{nte-eval-default}). This is in line with standard
practice in the related literature on CE (see e.g. Schut et al.
(\citeproc{ref-schut2021generating}{2021})). For our final results
presented in the main paper, we rely on held-out test sets for
evaluation. For tuning purposes we rely on training and/or validation
sets.

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]

@@

\bibitem[\citeproctext]{ref-schut2021generating}
Schut, Lisa, Oscar Key, Rory McGrath, Luca Costabello, Bogdan Sacaleanu,
Yarin Gal, et al. 2021. {``Generating {Interpretable Counterfactual
Explanations By Implicit Minimisation} of {Epistemic} and {Aleatoric
Uncertainties}.''} In \emph{International {Conference} on {Artificial
Intelligence} and {Statistics}}, 1756--64. {PMLR}.
